---
title: "Distribution_Analysis_1000"
author: "Anthony Song"
date: "3/26/2021"
output: html_document
---

```{r message=FALSE}
library(plyr)
library(tidyverse)
library(lubridate)
library(readxl)
library(ggplot2)
library(matrixStats)
library(dplyr)
library(xlsx)
library(glmnet)
library(ISLR)
library(dplyr)
library(caret)
library(ROCR)
library(class)
library(MASS)
library(randomForest)
#library(e1071)
```

### random forest multiclass cross validation. 
### Use multiclass_data.csv
## Returns a distribution of genes for each iteration. 
## Data must be as factor in order of the Random forrest to run. 

# 1.) 
```{r}
# Multiclass data
multiclass_dat <- read_csv('multiclass_data.csv')
multiclass_dat$multiclass = factor(multiclass_dat$multiclass) 
dat1<- multiclass_dat
dat1$X1 <- NULL
head(dat1)
```

```{r}
as.numeric(dat1)
```
## 2.)  MGMT methylated vs. MGMT unmethylated
```{r}
# MGMT methylated vs. MGMT unmethylated
MGMT_dat <- read_csv('MGMT_data.csv')
MGMT_dat$multiclass = factor(MGMT_dat$multiclass) 
dat2<- MGMT_dat
dat2$X1 <- NULL
head(dat2)
```
## 3.) IDH wt vs. IDH Mutant
```{r}

# IDH wildtype vs. IDH mutant
IDH_dat <- read_csv('IDH_data.csv')
IDH_dat$multiclass = factor(IDH_dat$multiclass) 
dat3<- IDH_dat
dat3$X1 <- NULL
head(dat3)

```
## 4.) MGMT only (IDHwt_MGMTm + IDHm_MGMTm = 139 samples) vs. Rest (IDHm_MGMTu+ IDHwt_MGMTu = 180 samples) 
```{r}
dat4 <- multiclass_dat
dat4 <- mutate(dat4 , multiclass = case_when(

  multiclass == "IDHwt_MGMTm"  ~ "MGMT_mutant",
  multiclass == "IDHm_MGMTm"  ~ "MGMT_mutant", 
  multiclass == "IDHm_MGMTu"  ~ "Rest",
  multiclass == "IDHwt_MGMTu"  ~ "Rest"
  
))
dat4$X1 <- NULL
dat4$multiclass = factor(dat4$multiclass) 
table(dat4$multiclass)
head(dat4)
```
## 5.)	IDHwt_MGMTu (173 samples) vs. the rest (IDHm_MGMTm + IDHm_MGMTu + IDHwt_MGMTm = 146 samples ) 
```{r}
dat5 <- multiclass_dat
dat5 <- mutate(dat5 , multiclass = case_when(

  multiclass == "IDHwt_MGMTu"  ~ "IDHwt_MGMTu",
  multiclass == "IDHm_MGMTm"  ~ "Rest", 
  multiclass == "IDHm_MGMTu"  ~ "Rest",
  multiclass == "IDHwt_MGMTm"  ~ "Rest"
  
))
dat5$X1 <- NULL
dat5$multiclass = factor(dat5$multiclass) 
table(dat5$multiclass)
head(dat5)
```


## 6.) IDH mutant by itself (IDHm_MGMTm + IDHm_MGMTu = 28 samples) vs Rest (IDHwt_MGMTm + IDHwt_MGMTu = 291 samples)
```{r}
dat6 <- multiclass_dat
dat6 <- mutate(dat6 , multiclass = case_when(

  multiclass == "IDHwt_MGMTu"  ~ "Rest",
  multiclass == "IDHm_MGMTm"  ~ "IDHm", 
  multiclass == "IDHm_MGMTu"  ~ "IDHm",
  multiclass == "IDHwt_MGMTm"  ~ "Rest"
  
))
dat6$X1 <- NULL
dat6$multiclass = factor(dat6$multiclass) 
head(dat6)
table(dat6$multiclass)
```

## 7.)	Inside MGMTu, can we distinguish between IDHm and IDHwt. Sample size = 180. (IDHm_MGMTu = 7 vs. IDHwt_MGMTu = 173)

```{r}
dat7 <- multiclass_dat
dat7 <- mutate(dat7 , multiclass = case_when(

  multiclass == "IDHwt_MGMTu"  ~ "IDHwt_insideMGMTu",
  multiclass == "IDHm_MGMTm"  ~ "rest", 
  multiclass == "IDHm_MGMTu"  ~ "IDHm_insideMGMTu",
  multiclass == "IDHwt_MGMTm"  ~ "rest"
  
))
dat7$X1 <- NULL
dat7<-dat7[!(dat7$multiclass=="rest"),]
dat7$multiclass = factor(dat7$multiclass) 
head(dat7)
table(dat7$multiclass)
```

## 8.)	Inside MGMTm, can we distingushi between IDHm and IDHwt. Sample size = 139. (IDHm_MGMTm = 21 and IDHwt_MGMTm = 118)
```{r}
dat8 <- multiclass_dat
dat8 <- mutate(dat8 , multiclass = case_when(

  multiclass == "IDHwt_MGMTu"  ~ "rest",
  multiclass == "IDHm_MGMTm"  ~ "IDHm_inside_MGMTm", 
  multiclass == "IDHm_MGMTu"  ~ "rest",
  multiclass == "IDHwt_MGMTm"  ~ "IDHwt_inside_MGMTm"
  
))
dat8$X1 <- NULL
dat8<-dat8[!(dat8$multiclass=="rest"),]
dat8$multiclass = factor(dat8$multiclass) 
head(dat8)
table(dat8$multiclass)
```


Check below later. 
## 9.) 	(IDHm_MGMTm + IDHwt_MGMTm = 139) vs. (IDHwt_MGMTm = 118) To see if the 21 IDH mutant samples  makes a difference at all. 

dat7 <- multiclass_dat
dat7 <- mutate(dat7 , multiclass = case_when(

  multiclass == "IDHwt_MGMTu"  ~ "",
  multiclass == "IDHm_MGMTm"  ~ "MGMTm_IDH", 
  multiclass == "IDHm_MGMTu"  ~ "",
  multiclass == "IDHwt_MGMTm"  ~ "MGMTm_IDH"
  
))


dat7<-dat7[!(dat7$multiclass=="B"),]
dat7$X1 <- NULL
head(dat7)
table(dat7$multiclass)





```{r}
table(multiclass_dat$multiclass)
```

```{r}

dat_test_1 = dat1[1:100, 1:100]
dat_test_1
```
#### This loop gets the distribution of the genes for 1000 CV loops. It is universal and can be used for all of them. 
```{r}

# This is the outer loop. We are interested in running it 1000 times to get a distribution of genes. 
n = 1000

for(i in 1:n){

# Need to start with a fresh data set everytime since we are decreasing in our inner loop. 
df <- dat1

cv_data <- data.frame(matrix(0, ncol = 3, nrow = 12))
names(cv_data)[1] <- "iteration"
names(cv_data)[2] <- "num_genes"
names(cv_data)[3] <- "prediction_accuracy"

optimal_gene_set <- c()

# cross validation loop. 
for(x in 1:10){

# name iteration number.   
cv_data[x, "iteration"] <- x

# partition training and test data. 
nindex <- createDataPartition(y = df$multiclass, p = 0.65, list = FALSE)

#set.seed(314)    # Set seed for reproducible results
#tindex <- sample(n, ntrain)   # Create a random index
train_df <- df[nindex,]   # Create training set
test_df <- df[-nindex,]   # Create test set


# create random forest object. 
modfit.rpart <- randomForest(multiclass ~ ., data=train_df, method="class", importance = TRUE)

# Predict the testing set with the trained model 
predictions1 <- predict(modfit.rpart, test_df, type = "class")

# Accuracy and other metrics
cm <- confusionMatrix(predictions1, test_df$multiclass)
cv_data[x, "prediction_accuracy"] <- cm$overall['Accuracy']

# Get genes by importance. 
ordered_genes <- data.frame(modfit.rpart$importance)

# order the genes by Mean Decrease in Accuracy. 
ordered_genes <- arrange(ordered_genes, desc(MeanDecreaseAccuracy))
ordered_genes

# check to see if the current prediction accuracy is better than the previous. 
#first iteration just set the first gene set to the optimal. 
if(x == 1){
  optimal_gene_set <- rownames(ordered_genes)
}
# else if current prediction accuracy is greater than or equal to the max accuracy then replace it. 
else if(cm$overall['Accuracy'] >= max(cv_data$prediction_accuracy, na.rm = TRUE)){
  optimal_gene_set <- rownames(ordered_genes)
}

# write number of ordered genes. 
cv_data[x, "num_genes"] <- nrow(ordered_genes)

# Take the top 50% of genes with the highest mean decrease in accuracy. 
half_genes <- ordered_genes[1:ceiling(0.5 * nrow(ordered_genes)),]

# replace data frame with top 50% genes.  
df <- df[ prepend(rownames(half_genes), "multiclass")] 

# reiterate. 
}

# In the first iteration we create the table. 
if(i == 1){
table <- as.data.frame(table(optimal_gene_set))
}
# else we add to the table. 
else{
  
# This loop goes through the frequency data frame and adds the current vector of genes to the dataframe. 
for (gene in optimal_gene_set){
  
  # if the gene is in the current table
  if( any(gene %in% table$optimal_gene_set)){
    
    #get the index of the gene in the data frame. 
    index <- which(grepl(gene, table$optimal_gene_set))
    # increase the frequency by 1. 
    table[index,2] <- table[index,2]+1
    
  }
  
  # if gene is not in the current gene set add it to the gene set with frequency of 1
  else{
    
    # create a data frame with the new gene to add. 
    temp<-data.frame(gene,1)
    names(temp)<-c(colnames(table)[1], colnames(table)[2])

    # bind the two data frames. 
    table <- rbind(table, temp)
    
    
  }
}
  
}

}
multiclass_table <- table
write.csv(multiclass_table,"multiclass_gene_distribution.csv")
```
```{r}
table
```

### Writes the table to .csv
```{r}
multiclass_table <- table
write.csv(multiclass_table,"multiclass_gene_distribution.csv")
```