---
title: "multi_classification_MGMT_IDH"
author: "Anthony Song"
date: "12/16/2020"
output: html_document
---



```{r message=FALSE}
library(plyr)
library(xgboost)
library(e1071)
library(tidyverse)
library(lubridate)
library(ggpubr)
library(readxl)
library(ggplot2)
library(ggpubr)
library(matrixStats)
library(dplyr)
library(xlsx)
library(NOISeq)
library(glmnet)
library(ISLR)
library(dplyr)
library(caret)
library(ROCR)
library(class)
library(MASS)
library(randomForest)
```


### Read in Pheno and g exp data. 
```{r message = FALSE}

# read in patient clinical information for Multiclass  
info_og <- read_csv('TCGA_GBM_HT_HG-U133A_patient_information_v1.csv')
info <- info_og

g_exp_og <- read_csv('TCGA_GBM_HT_HG-U133A_geneEXP_v2.csv')
# rename column 1
names(g_exp_1)[1] <- "gene"
```

### MULTICLASS PIPELINE: Clean the data into a usable data frame. 
```{r, message = FALSE}

### For subtype IDH Mutation Status  
info <- info[!is.na(info$subtype_IDH.status),]
### For subtype MGMT promoter status. 
info <- info[!is.na(info$subtype_MGMT.promoter.status),]

# Select data with only tumor GBM patients. 
info <-info %>%
  filter(shortLetterCode == "TP")

# Get relevant columns
columns <- c("barcode", "subtype_IDH.status", "subtype_MGMT.promoter.status")
info <- info[columns]


# Split into IDHwt MGMTm, IDHwt MGMu, IDHm MGMTm, and IDHm MGMTu classes
# Multiclass column Legend: 
# 1 = IDHwt MGMTm
# 2 = IDHwt MGMu
# 3 = IDHm MGMTm
# 4 = IDHm MGMTu

info <- mutate(info , multiclass = case_when(
  subtype_IDH.status == "WT" & subtype_MGMT.promoter.status == "Methylated" ~ "IDHwt_MGMTm",
  subtype_IDH.status == "WT" & subtype_MGMT.promoter.status == "Unmethylated" ~ "IDHwt_MGMTu",
  subtype_IDH.status == "Mutant" & subtype_MGMT.promoter.status == "Methylated" ~ "IDHm_MGMTm", 
  subtype_IDH.status == "Mutant" & subtype_MGMT.promoter.status == "Unmethylated" ~ "IDHm_MGMTu"
))

# Get frequency
table(info$multiclass)

# Use all genes. 


require(MASS)
require(dplyr)

# get all sample names
source.names <- c( "gene" ,info$barcode)
length(source.names)

g_exp_1 <- g_exp_og
colnames(g_exp_1)[1] = "gene"

# filter expression data for sample names, and get distinct rows.  
g_exp_1 <- g_exp_1 %>%
  dplyr::select(source.names) %>%
  mutate(gene= factor(gene)) %>%
  distinct(gene,  .keep_all = TRUE)
ncol(g_exp_1)# 320
nrow(g_exp_1) # 10862

# Remove genes with a dash in the name. Because it is not compatible with the random forrest. 
g_exp_1 <- g_exp_1[!grepl("-", g_exp_1$gene),]
nrow(g_exp_1) # 10829

head(g_exp_1)

# Get gene expression data with only the genes of interest. 
g_exp_temp <- g_exp_1

# transpose data frame. 
g_exp_temp <- as.data.frame(t(as.matrix(g_exp_temp)))
# turn first row into column names. 
colnames(g_exp_temp) <- as.character(unlist(g_exp_temp[1,]))
g_exp_temp = g_exp_temp[-1, ]
# Convert all columns into numeric values. 
g_exp_temp[] <- lapply(g_exp_temp, function(x) as.numeric(as.character(x)))

# Merge into a final dataframe. 
df <- cbind(info$multiclass, g_exp_temp)

# Multiclass column Legend: 
# 1 = IDHwt MGMTm
# 2 = IDHwt MGMu
# 3 = IDHm MGMTm
# 4 = IDHm MGMTu
names(df)[names(df) == "info$multiclass"] <- "multiclass"
df$multiclass <- as.factor(df$multiclass)
head(df)
dat <- df
```

# Write to .csv to current directory
```{r}
write.csv(dat,"multiclass_data.csv")
```


# create small data set for testing
```{r}
test_dat <- dat[1:100,1:100]
test_dat
```
### random forest multiclass cross validation. 
## Returns a distribution of genes for each iteration. 
```{r}

# This is the outer loop. We are interested in running it 1000 times to get a distribution of genes. 
n = 100

for(i in 1:n){

# Need to start with a fresh data set everytime since we are decreasing in our inner loop. 
df <- test_dat

cv_data <- data.frame(matrix(0, ncol = 3, nrow = 12))
names(cv_data)[1] <- "iteration"
names(cv_data)[2] <- "num_genes"
names(cv_data)[3] <- "prediction_accuracy"

optimal_gene_set <- c()

# cross validation loop. 
for(x in 1:10){

# name iteration number.   
cv_data[x, "iteration"] <- x

# partition training and test data. 
nindex <- createDataPartition(y = df$multiclass, p = 0.65, list = FALSE)

#set.seed(314)    # Set seed for reproducible results
#tindex <- sample(n, ntrain)   # Create a random index
train_df <- df[nindex,]   # Create training set
test_df <- df[-nindex,]   # Create test set


# create random forest object. 
modfit.rpart <- randomForest(multiclass ~ ., data=train_df, method="class", importance = TRUE)

# Predict the testing set with the trained model 
predictions1 <- predict(modfit.rpart, test_df, type = "class")

# Accuracy and other metrics
cm <- confusionMatrix(predictions1, test_df$multiclass)
cv_data[x, "prediction_accuracy"] <- cm$overall['Accuracy']

# Get genes by importance. 
ordered_genes <- data.frame(modfit.rpart$importance)

# order the genes by Mean Decrease in Accuracy. 
ordered_genes <- arrange(ordered_genes, desc(MeanDecreaseAccuracy))
ordered_genes

# check to see if the current prediction accuracy is better than the previous. 
#first iteration just set the first gene set to the optimal. 
if(x == 1){
  optimal_gene_set <- rownames(ordered_genes)
}
# else if current prediction accuracy is greater than or equal to the max accuracy then replace it. 
else if(cm$overall['Accuracy'] >= max(cv_data$prediction_accuracy, na.rm = TRUE)){
  optimal_gene_set <- rownames(ordered_genes)
}

# write number of ordered genes. 
cv_data[x, "num_genes"] <- nrow(ordered_genes)

# Take the top 50% of genes with the highest mean decrease in accuracy. 
half_genes <- ordered_genes[1:ceiling(0.5 * nrow(ordered_genes)),]

# replace data frame with top 50% genes.  
df <- df[ prepend(rownames(half_genes), "multiclass")] 

# reiterate. 
}

# In the first iteration we create the table. 
if(i == 1){
table <- as.data.frame(table(optimal_gene_set))
}
# else we add to the table. 
else{
  
# This loop goes through the frequency data frame and adds the current vector of genes to the dataframe. 
for (gene in optimal_gene_set){
  
  # if the gene is in the current table
  if( any(gene %in% table$optimal_gene_set)){
    
    #get the index of the gene in the data frame. 
    index <- which(grepl(gene, table$optimal_gene_set))
    # increase the frequency by 1. 
    table[index,2] <- table[index,2]+1
    
  }
  
  # if gene is not in the current gene set add it to the gene set with frequency of 1
  else{
    
    # create a data frame with the new gene to add. 
    temp<-data.frame(gene,1)
    names(temp)<-c(colnames(table)[1], colnames(table)[2])

    # bind the two data frames. 
    table <- rbind(table, temp)
    
    
  }
}
  
}

}

table
cv_data

```
```{r}
colnames(table)

```

# This is the architecture of the loop to create the dictionary data frame. 
```{r}
table <- as.data.frame(table(multiclass_genes))

test <- c("CBR1", "MGMT", "FBXO17", "DDB2", "YOLO69", "Paster")


# This loop goes through the frequency data frame and adds the current vector of genes to the dataframe. 
for (gene in test){
  
  # if the gene is in the current table
  if( any(gene %in% table$multiclass_genes)){
    
    #get the index of the gene in the data frame. 
    index <- which(grepl(gene, table$multiclass_genes))
    # increase the frequency by 1. 
    table[index,2] <- table[index,2]+1
    
  }
  
  # if gene is not in the current gene set add it to the gene set with frequency of 1
  else{
    
    # create a data frame with the new gene to add. 
    temp<-data.frame(gene,1)
    names(temp)<-c("multiclass_genes","Freq")

    # bind the two data frames. 
    table <- rbind(table, temp)
    
    
  }
}
table

```

# Information on Optimal performance for multiclass setting. 
```{r}
multiclass_genes <- rownames(ordered_genes)
multiclass_genes
cm
```






### DICHOTOMOUS MGMT PIPELINE: Clean the data into a usable data frame. 
```{r, message = FALSE}
# for dichotomous MGMT
info1 <- info_og

### For subtype MGMT promoter status. 
info1 <- info1[!is.na(info1$subtype_MGMT.promoter.status),]

# Select data with only tumor GBM patients. 
info1 <-info1 %>%
  filter(shortLetterCode == "TP")

# Get relevant columns
columns <- c("barcode", "subtype_IDH.status", "subtype_MGMT.promoter.status")
info1 <- info1[columns]


# Split into IDHwt MGMTm, IDHwt MGMu, IDHm MGMTm, and IDHm MGMTu classes
# Multiclass column Legend: 
# 1 = MGMTm
# 2 = MGMu

info1 <- mutate(info1 , multiclass = case_when(
  subtype_MGMT.promoter.status == "Methylated" ~ "MGMTm",
  subtype_MGMT.promoter.status == "Unmethylated" ~ "MGMTu"
))

# Get frequency
table(info1$multiclass)
```

```{r}
require(MASS)
require(dplyr)

# get all sample names
source.names <- c( "gene" ,info1$barcode)
length(source.names)

g_exp_2 <- g_exp_og
colnames(g_exp_2)[1] = "gene"

# filter expression data for sample names, and get distinct rows.  
g_exp_2 <- g_exp_2 %>%
  dplyr::select(source.names) %>%
  mutate(gene= factor(gene)) %>%
  distinct(gene,  .keep_all = TRUE)
ncol(g_exp_2)# 364
nrow(g_exp_2) # 10862

# Remove genes with a dash in the name. Because it is not compatible with the random forrest. 
g_exp_2 <- g_exp_2[!grepl("-", g_exp_2$gene),]
nrow(g_exp_2) # 10829

head(g_exp_2)

# Get gene expression data with only the genes of interest. 
g_exp_temp <- g_exp_2

# transpose data frame. 
g_exp_temp <- as.data.frame(t(as.matrix(g_exp_temp)))
# turn first row into column names. 
colnames(g_exp_temp) <- as.character(unlist(g_exp_temp[1,]))
g_exp_temp = g_exp_temp[-1, ]
# Convert all columns into numeric values. 
g_exp_temp[] <- lapply(g_exp_temp, function(x) as.numeric(as.character(x)))

# Merge into a final dataframe. 
df <- cbind(info1$multiclass, g_exp_temp)

names(df)[names(df) == "info1$multiclass"] <- "multiclass"
df$multiclass <- as.factor(df$multiclass)
head(df)
```
# Write to .csv to current directory
```{r}
write.csv(df,"MGMT_data.csv")
```

### random forest MGMT dichotomous classification cross validation. 
#MGMT dichotomous classification 
```{r}
# create data frame to fill in. Will run 12 iterations.
cv_data1 <- data.frame(matrix(0, ncol = 3, nrow = 12))
names(cv_data1)[1] <- "iteration"
names(cv_data1)[2] <- "num_genes"
names(cv_data1)[3] <- "prediction_accuracy"



# MGMT dichotomous cross validation loop. 

for(x in 1:9){

# name iteration number.   
cv_data1[x, "iteration"] <- x

# partition training and test data. 
nindex <- createDataPartition(y = df1$multiclass, p = 0.65, list = FALSE)

#set.seed(314)    # Set seed for reproducible results
#tindex <- sample(n, ntrain)   # Create a random index
train_df1 <- df1[nindex,]   # Create training set
test_df1 <- df1[-nindex,]   # Create test set


# create random forest object. 
modfit.rpart <- randomForest(multiclass ~ ., data=train_df1, method="class", importance = TRUE)

# Predict the testing set with the trained model 
predictions2 <- predict(modfit.rpart, test_df1, type = "class")

# Accuracy and other metrics
cm1 <- confusionMatrix(predictions2, test_df1$multiclass)
cv_data1[x, "prediction_accuracy"] <- cm1$overall['Accuracy']

# Get genes by importance. 
ordered_genes_mgmt <- data.frame(modfit.rpart$importance)

# order the genes by Mean Decrease in Accuracy. 
ordered_genes_mgmt <- arrange(ordered_genes_mgmt, desc(MeanDecreaseAccuracy))
ordered_genes_mgmt

# write number of ordered genes. 
cv_data1[x, "num_genes"] <- nrow(ordered_genes_mgmt)

# Take the top 50% of genes with the highest mean decrease in accuracy. 
half_genes_mgmt <- ordered_genes_mgmt[1:ceiling(0.5 * nrow(ordered_genes_mgmt)),]

# replace data frame with top 50% genes.  
df1 <- df1[ prepend(rownames(half_genes_mgmt), "multiclass")] 

# reiterate. 
}


cv_data1

```

```{r}
mgmt_optimal_MDA_genes <- rownames(ordered_genes_mgmt)
mgmt_optimal_MDA_genes
#varImpPlot(modfit.rpart)
#cm1
```


### DICHOTOMOUS IDH PIPELINE: Clean the data into a usable data frame. 

```{r, message = FALSE}
# for dichotomous IDH
info2 <- info_og

### For subtype IDH promoter status. 
info2 <- info2[!is.na(info2$subtype_IDH.status),]

# Select data with only tumor GBM patients. 
info2 <-info2 %>%
  filter(shortLetterCode == "TP")

# Get relevant columns
columns <- c("barcode", "subtype_IDH.status", "subtype_MGMT.promoter.status")
info2 <- info2[columns]


# Split into IDHwt MGMTm, IDHwt MGMu, IDHm MGMTm, and IDHm MGMTu classes
# Multiclass column Legend: 
# 1 = IDHm
# 2 = IDHwt

info2 <- mutate(info2 , multiclass = case_when(

  subtype_IDH.status == "WT"  ~ "IDHwt",
  subtype_IDH.status == "Mutant"  ~ "IDHm"
  
))

# Get frequency
table(info2$multiclass)
```

```{r}
head(info2)
```

```{r}
require(MASS)
require(dplyr)

# get all sample names
source.names <- c( "gene" ,info2$barcode)
length(source.names)

g_exp_3 <- g_exp_og
colnames(g_exp_3)[1] = "gene"

# filter expression data for sample names, and get distinct rows.  
g_exp_3 <- g_exp_3 %>%
  dplyr::select(source.names) %>%
  mutate(gene= factor(gene)) %>%
  distinct(gene,  .keep_all = TRUE)
ncol(g_exp_3)# 434
nrow(g_exp_3) # 10862

# Remove genes with a dash in the name. Because it is not compatible with the random forrest. 
g_exp_3 <- g_exp_3[!grepl("-", g_exp_3$gene),]
nrow(g_exp_3) # 10829

head(g_exp_3)

# Get gene expression data with only the genes of interest. 
g_exp_temp <- g_exp_3

# transpose data frame. 
g_exp_temp <- as.data.frame(t(as.matrix(g_exp_temp)))
# turn first row into column names. 
colnames(g_exp_temp) <- as.character(unlist(g_exp_temp[1,]))
g_exp_temp = g_exp_temp[-1, ]
# Convert all columns into numeric values. 
g_exp_temp[] <- lapply(g_exp_temp, function(x) as.numeric(as.character(x)))

# Merge into a final dataframe. 
df <- cbind(info2$multiclass, g_exp_temp)

names(df)[names(df) == "info2$multiclass"] <- "multiclass"
df$multiclass <- as.factor(df$multiclass)
head(df)
```
# Write to .csv to current directory
```{r}
write.csv(df,"IDH_data.csv")
```


### random forest IDH dichotomous classification cross validation. 

#IDH dichotomous classification 
```{r}
# create data frame to fill in. Will run 12 iterations.
cv_data2 <- data.frame(matrix(0, ncol = 3, nrow = 12))
names(cv_data2)[1] <- "iteration"
names(cv_data2)[2] <- "num_genes"
names(cv_data2)[3] <- "prediction_accuracy"


# MGMT dichotomous cross validation loop. 

for(x in 1:9){

# name iteration number.   
cv_data2[x, "iteration"] <- x

# partition training and test data. 
nindex <- createDataPartition(y = df2$multiclass, p = 0.65, list = FALSE)

#set.seed(314)    # Set seed for reproducible results
#tindex <- sample(n, ntrain)   # Create a random index
train_df2 <- df2[nindex,]   # Create training set
test_df2 <- df2[-nindex,]   # Create test set


# create random forest object. 
modfit.rpart <- randomForest(multiclass ~ ., data=train_df2, method="class", importance = TRUE)

# Predict the testing set with the trained model 
predictions3 <- predict(modfit.rpart, test_df2, type = "class")

# Accuracy and other metrics
cm2 <- confusionMatrix(predictions3, test_df2$multiclass)
cv_data2[x, "prediction_accuracy"] <- cm2$overall['Accuracy']

# Get genes by importance. 
ordered_genes_idh <- data.frame(modfit.rpart$importance)

# order the genes by Mean Decrease in Accuracy. 
ordered_genes_idh <- arrange(ordered_genes_idh, desc(MeanDecreaseAccuracy))
ordered_genes_idh

# write number of ordered genes. 
cv_data2[x, "num_genes"] <- nrow(ordered_genes_idh)

# Take the top 50% of genes with the highest mean decrease in accuracy. 
half_genes_idh <- ordered_genes_idh[1:ceiling(0.5 * nrow(ordered_genes_idh)),]

# replace data frame with top 50% genes.  
df2 <- df2[ prepend(rownames(half_genes_idh), "multiclass")] 

# reiterate. 
}


cv_data2

```

```{r}
idh_optimal_MDA_genes <- rownames(ordered_genes_idh)
ordered_genes_idh
varImpPlot(modfit.rpart)
cm2
```

```{r}
idh_optimal_MDA_genes 
mgmt_optimal_MDA_genes
multiclass_genes
```
```{r}
idh_mgmt_intersect <- intersect(idh_optimal_MDA_genes ,mgmt_optimal_MDA_genes )
idh_mgmt_intersect

idh_multiclass_intersect <- intersect(idh_optimal_MDA_genes ,multiclass_genes )
idh_multiclass_intersect

mgmt_multiclass_intersect <- intersect(mgmt_optimal_MDA_genes, multiclass_genes)
mgmt_multiclass_intersect 

idh_mgmt_union_1 <- union(idh_optimal_MDA_genes ,mgmt_optimal_MDA_genes )
set_diff_multiclass_1 <-setdiff(multiclass_genes, idh_mgmt_union_1)
set_diff_multiclass_1
```

# PPT Intersections: 
```{r}
multiclass_genes_1 <- c("MGMT", "DIS3", "FBXO17", "IFNA21", "DYNLT3", "CBR1", "BGH1", "EMP3", "DRG2", "RBP1", "TIMP1", "HTATIP2", "TRIP4", "TAGLN2", "DDB2", "PER1", "FCHSD2", "RARRES2", "MSN", "CRIPT", "CXCL9", "TMEM100")

mgmt_genes <- c("MGMT", "TPSG1", "DDB2", "ANXA7", "CYP46A1", "PAXIP1", "HTATIP2", "HIRA", "FIG4", "RRAGD", "HEY1", "CCNC", "HSPBP1", "CCDC121", "SLC2A10", "OBSL1", "USP32", "GPI", "KPNA5", "NSUN5", "NOX1", "TRMT11", "CUEDC2", "ACTN4", "HCCS", "INPP5E", "HYOU1", "ABCG5", "RNF146", "SLC4A8", "S100A13", "ITGA4", "GIPC1", "TJAP1", "UXS1", "HIST1H31", "LRP3", "EXOSC4", "ADM", "ASNA1", "ABCG2", "ZBTB16", "ZAP70")


idh_genes <- c("FBXO17", "CBR1", "TIMP1", "TAGLN2", "RBP1", "DDB2", "EMP3", "TRIP4", "C1QL1", "RAB36", "FHL2", "BMP2", "WWTR1", "EFEMP2", "ZNF248", "DRG2", "TMEM100", "PNPLA4", "PHLDA3", "MEOX2", "SLC25A20", "PLP2", "SLC43A3", "BDH1", "ATXN7L1", "AKAP6",  "ABI1", "ALDOA", "LRRC49", "SLC27A3", "ALCAM", "BMP5", "PRPS2", "DAZ4", "CD97", "GDF15", "CRTAC1", "PDGFA", "SEC61G", "FRY", "AMFR", "SEC31B", "ARHGAP12")

idh_mgmt_intersection <- intersect(idh_genes ,mgmt_genes )
idh_mgmt_intersection 


multiclass_idh_intersect_1 <- intersect(multiclass_genes_1, idh_genes)
multiclass_idh_intersect_1

multiclass_mgmt_intersect_1 <- intersect(multiclass_genes_1, mgmt_genes)
multiclass_mgmt_intersect_1

idh_mgmt_union <- union(idh_genes ,mgmt_genes )
set_diff_multiclass <-setdiff(multiclass_genes_1, idh_mgmt_union )
set_diff_multiclass
```




